---
title: "hw6_nonames2"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/maxand22/Google Drive/Humboldt/4. Semester/Machine Intelligence II/Machine_Intelligence_II/")
knitr::opts_chunk$set(error = TRUE)
#load("test.RData")
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

#https://stackoverflow.com/questions/11155182/is-there-a-way-to-knitr-markdown-straight-out-of-your-workspace-using-rstudio
```

```{r results='hide', message=FALSE, warning=FALSE}
library(audio)

#install.packages("VGAM")
library(VGAM)

#install.packages("R.matlab")
library(R.matlab)
library(ggplot2)
```


6.1.a 
Extend your code from the previous problem sheet to get an ICA-learning scheme based on the natural gradient with a learning rate ε that decays slowly to 0 (e.g. εt+1 = λεt with λ ≈ 1, λ < 1). Note that depending on λ you have to iterate over the (shuffled) data more than once for proper convergence.

Initializsation of last Excersise
```{r}
# a)
s1 = read.table("hw5/sound1.dat", header=FALSE)
s2 = read.table("hw5/sound2.dat", header=FALSE)

S = t(as.matrix(data.frame(s1,s2)))

# b)
set.seed(1234)
A = matrix(runif(4, 0, 1), nrow = 2)

X = A%*%S


# e)
X[1,] = X[1,] - mean(X[1,])
X[2,] = X[2,] - mean(X[2,])
```


Extention by lambda decreasing to zero and Plot of S_head
(assumption lambda ist not the eigenvalue)
```{r}
# logistic function
f = function(x){
    return(1/(1+exp(-x)))
}


# natural Gradient
t = 1
eta_t = 0.7
alpha = 1
lambda = 0.9102 #0.9112
set.seed(9991)
W2 = matrix(runif(4, 0, 1), ncol = 2)


unmixing_natural = function(W, X, n_steps = 18000){
    
    for(t in 1:n_steps){
        eta_t = eta_t*lambda
        
        if(alpha > n_steps-12){
           print(eta_t)
        }
        
        x = X[,alpha]
        
        c <-  do.call(cbind, replicate(nrow(W), x, simplify=FALSE))
        
        f_wx = 1 - 2*f(W%*%c)
        
        wx = W%*%c
        
        k_delta = diag(nrow(W))
        
        
        W_delta = eta_t*((k_delta + f_wx%*%wx)%*%W)
        
        W = W + W_delta
        
        alpha = alpha + 1
        if(alpha == n_steps){
            alpha = 1
        }
    }
    return(W)
}


W_natural = unmixing_natural(W2, X)
S_hat_natural = W_natural%*%X

cor(t(S_hat_natural), t(S))
par(mfrow = c(1,2))
plot(1:18000, S_hat_natural[1,], type = 'l', main = 'unmixed bird', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_natural[2,], type = 'l', main = 'unmixed halleluja', xlab = 't', ylab = 'value')

#play(audioSample(t(as.matrix(S_hat_natural[1,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural[2,])), rate = 8192))

```


6.1.b
Use the two sound signals from the last problem sheet and add (as third source s3) an ad- ditional “noise” source (normally distributed random numbers with a standard deviation similar to the two signals). Mix the signals using a mixing matrix of your choice and apply your ICA-algorithm. Plot the Mixed Sounds and recovered Sources


Plot initialised the third source normal distributed
```{r}

set.seed(222)
s3 <- t(matrix(rnorm(nrow(s1),0,((sd(as.matrix(s1)) + sd(as.matrix(s2)))/2)),nrow = 1))
S3 = t(as.matrix(data.frame(s1,s2,s3)))

par(mfrow=c(1,1))
plot(1:18000, s3, type = 'l', main = 'source 3 normal distributed', xlab = 't', ylab = 'value')

```

plot the created observations X
```{r}

set.seed(1234)
A3 = matrix(runif(9, 0, 1), nrow = 3)
X3 = A3%*%S3

#centering
X3[1,] = X3[1,] - mean(X3[1,])
X3[2,] = X3[2,] - mean(X3[2,])
X3[3,] = X3[3,] - mean(X3[3,])

par(mfrow = c(2,2))
plot(1:18000, X3[1,], type = 'l', main = 'observation 1', xlab = 't', ylab = 'value')
plot(1:18000, X3[2,], type = 'l', main = 'observation 2', xlab = 't', ylab = 'value')
plot(1:18000, X3[3,], type = 'l', main = 'observation 3', xlab = 't', ylab = 'value')


```


Plot the unmixed Xb by modified natural gradient 3x3
```{r}
# natural Gradient
t = 1
eta_t = 0.3
alpha = 1
lambda = 0.9102 #0.9102
set.seed(9991)
W3 = matrix(runif(9, 0, 1), ncol = 3)

W_natural3 = unmixing_natural(W3, X3)
S_hat_natural3 = W_natural3%*%X3

cor(t(S_hat_natural3), t(S3))

par(mfrow=c(2,2))
plot(1:18000, S_hat_natural3[1,], type = 'l', main = 's_hat bird', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_natural3[2,], type = 'l', main = 's_hat halleluja', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_natural3[3,], type = 'l', main = 's_hat source 3 normal distributed', xlab = 't', ylab = 'value')
#play(audioSample(t(as.matrix(S_hat_natural3[1,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[2,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[3,])), rate = 8192))
```


(c) Do the same analysis but adding a different “noise”-source (e.g. Laplace distributed) instead of the normal one.

Plot initialised the third source t distributed
```{r}

set.seed(666666)
st <- t(matrix(rt(nrow(s1),10),nrow = 1))
St = t(as.matrix(data.frame(s1,s2,st)))

par(mfrow=c(1,1))
plot(1:18000, st, type = 'l', main = 'source 3 t distributed', xlab = 't', ylab = 'value')

```

plot the created observations X laplace distributed
```{r}

set.seed(1234)
A3 = matrix(runif(9, 0, 1), nrow = 3)
Xt = A3%*%St

Xt[1,] = Xt[1,] - mean(Xt[1,])
Xt[2,] = Xt[2,] - mean(Xt[2,])
Xt[3,] = Xt[3,] - mean(Xt[3,])

par(mfrow=c(2,2))
plot(1:18000, Xt[1,], type = 'l', main = 'observation 1', xlab = 't', ylab = 'value')
plot(1:18000, Xt[2,], type = 'l', main = 'observation 2', xlab = 't', ylab = 'value')
plot(1:18000, Xt[3,], type = 'l', main = 'observation 3', xlab = 't', ylab = 'value')


```

plot S_hat noise source laplace distrubuted
```{r}
# natural Gradient
t = 1
eta_t = 0.25
alpha = 1
lambda = 0.91 #0.9102
set.seed(9991)
Wt = matrix(runif(9, 0, 1), ncol = 3)

W_naturalt = unmixing_natural(Wt, Xt)
S_hat_naturalt = W_naturalt%*%Xt

cor(t(S_hat_natural3), t(S3))

par(mfrow=c(2,2))
plot(1:18000, S_hat_naturalt[1,], type = 'l', main = 's_hat bird', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_naturalt[2,], type = 'l', main = 's_hat halleluja', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_naturalt[3,], type = 'l', main = 's_hat source 3 t distributed', xlab = 't', ylab = 'value')
#play(audioSample(t(as.matrix(S_hat_natural3[1,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[2,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[3,])), rate = 8192))
```




6.2 Moments of univariate distributions

Calculate the first 4 moments of the different random variables depending on the respective pa- rameters. In addition to providing the derivation (e.g. by using the characteristic function) fill the following table:



initalize the sample Laplace, Normal and Uniform Distribution
```{r}
M = matrix(1, ncol = 3, nrow = 4)

set.seed(14321)
#Laplace
b <- 1
mean <- 0
S_l <- rlaplace(nrow(s1),mean,b)

#normal
s3 <- rnorm(nrow(s1),0,((sd(as.matrix(s1)) + sd(as.matrix(s2)))/2))

#Uniform
a = 0
b = 1
S_uni <- runif(nrow(s1),a,b)

par(mfrow=c(2,2))
hist(S_l, main = "Histogram of Laplace Sample")
hist(s3, main = "Historgram of Normal Distribution")
hist(S_uni, main = "Histogram of Uniform Distribution")
```


calculate the first Moment
```{r}
M = matrix(1, ncol = 3, nrow = 4)

first_moment <- function(x){
  return(sum(x)/(length(x)))
}

M[1,1] <- round(first_moment(S_l),1)
M[1,2] <- round(first_moment(s3),1)
M[1,3] <- round(first_moment(S_uni),1)

M[1,]
```

calculate the second Moment
```{r}

second_moment <- function(x){
  return(sum((x-mean(x))^2)/length(x))
}

M[2,1] <- round(second_moment(S_l),1)
M[2,2] <- round(second_moment(s3),1)
M[2,3] <- round(second_moment(S_uni),1)

M[2,]
```


calculate the third Moment
```{r}
third_moment <- function(x){
  return((sum((x-mean(x))^3)/length(x))/((sum((x-mean(x))^2)/length(x))^(3/2)))
}

M[3,1] <- round(third_moment(S_l),1)
M[3,2] <- round(third_moment(s3),1)
M[3,3] <- round(third_moment(S_uni),1)

M[3,]
```


calculate the fourth Moment
```{r}
vier_moment <- function(x){
  return((sum((x-mean(x))^4)/length(x))/((sum((x-mean(x))^2)/length(x))^(4/2)))
}

M[4,1] <- round(vier_moment(S_l),1)
M[4,2] <- round(vier_moment(s3),1)
M[4,3] <- round(vier_moment(S_uni),1)

M[4,]
```

fill the following table: CHECK

the colums represent the distributions (Laplace, Normal, Uniform)

the row represent the Moments, the first, the second (centered), the third (standardized), the fourth(standardized)
```{r}
M
```



6.3
The file distrib.mat contains three toy datasets (uniform, normal, laplacian), each 10000 samples of 2 sources. Do the following for each dataset (which can be read for example using Python with loadmat from scipy.io):

```{r}
l = readMat("hw6/distrib.mat")

s1 = l[[1]]
s2 = l[[2]]
s3 = l[[3]]

```


(a) Apply the following mixing matrix A to the original data s:
```{r}
A = matrix(4:1, ncol = 2, byrow = TRUE)

x1 = A%*%s1
x2 = A%*%s2
x3 = A%*%s3

```


(b) Center the mixed data to zero mean.
```{r}
x1[1,] = x1[1,] - mean(x1[1,])
x1[2,] = x1[2,] - mean(x1[2,])

x2[1,] = x2[1,] - mean(x2[1,])
x2[2,] = x2[2,] - mean(x2[2,])

x3[1,] = x3[1,] - mean(x3[1,])
x3[2,] = x3[2,] - mean(x3[2,])

```


(c) Decorrelatethedatabyapplyingprincipalcomponentanalysis(PCA)andprojectthemonto the principal components (PCs).
```{r}
pcs1 = eigen(cov(t(x1)))$vectors
pcs2 = eigen(cov(t(x2)))$vectors
pcs3 = eigen(cov(t(x3)))$vectors

evals1 = eigen(cov(t(x1)))$values
evals2 = eigen(cov(t(x2)))$values
evals3 = eigen(cov(t(x3)))$values

proj1 = t(x1) %*% pcs1
proj2 = t(x2) %*% pcs2
proj3 = t(x3) %*% pcs3

round(cov((proj1)), 5)
round(cov((proj2)), 5)
round(cov((proj3)), 5)
```


(d) Scale the data to unit variance in each PC direction (now the data is whitened or sphered).
```{r}
proj1_w = t(x1) %*% pcs1 %*% diag(evals1^(-.5))
proj2_w = t(x2) %*% pcs2 %*% diag(evals2^(-.5))
proj3_w = t(x3) %*% pcs3 %*% diag(evals3^(-.5))

round(cov((proj1_w)), 5)
round(cov((proj2_w)), 5)
round(cov((proj3_w)), 5)
```


(e) Rotate the data by different angles θ and calculate the kurtosis1 empirically for each dimension:
```{r}
rotation_matrix = function(theta){
    return(matrix(c(cos(theta), -sin(theta), sin(theta), cos(theta)), ncol = 2, byrow = TRUE))
}

thetas = seq(0, 2*pi, pi/50)

kurt = function(x){
    k = 1/length(x)*sum(x^4) - 3
    return(k)
}

kurt_rotation = function(x, thetas){
    
    k_mat = matrix(nrow = length(thetas), ncol = 2)
    i = 1
    
    if(nrow(x) > ncol(x)){
        x = t(x)
    }
    
    for(t in thetas){
        R = rotation_matrix(t)
        
        x_theta = t(R%*%x)
        k = apply(x_theta, 2, kurt)
        
        k_mat[i,] = k
        i = i + 1
        
    }
    return(k_mat)
}

k_after_rotation1 = kurt_rotation(proj1_w, thetas)
k_after_rotation2 = kurt_rotation(proj2_w, thetas)
k_after_rotation3 = kurt_rotation(proj3_w, thetas)
```


(f) Find the minimum and maximum kurtosis value for the first dimension and rotate the data accordingly.
```{r}
max1 = which.max(k_after_rotation1[,1])
max2 = which.max(k_after_rotation2[,1])
max3 = which.max(k_after_rotation3[,1])

R1 = rotation_matrix(thetas[max1])
R2 = rotation_matrix(thetas[max2])
R3 = rotation_matrix(thetas[max3])

proj1_w_rotated = R1 %*% t(proj1_w)
proj2_w_rotated = R2 %*% t(proj2_w)
proj3_w_rotated = R3 %*% t(proj3_w)

```

• Plot the original dataset (sources) and the mixed dataset after the steps (a), (b), (c), (d), and (f) as a scatter plot and display the respective marginal histograms. For step (e) plot the kurtosis value as a function of angle for each dimension.
```{r}
m <- rbind(c(1, 2), c(1, 3))
layout(m, widths = 1)

plot(s1[1,], s1[2,], col = alpha("blue", .3), pch = 16,
     main = "Normal (original sources)", xlab = "s1", ylab = "s2")
hist(s1[1,], main = "Normal s1 original", xlab = "")
hist(s1[2,], main = "Normal s2 original", xlab = "")

plot(proj1_w_rotated[1,], proj1_w_rotated[2,], col = alpha("blue", .3), pch = 16,
     main = "Normal (whitened & rotated)", xlab = "s1", ylab = "s2")
hist(proj1_w_rotated[1,], main = "Normal s1 whitened & rotated", xlab = "")
hist(proj1_w_rotated[2,], main = "Normal s2 whitened & rotated", xlab = "")


plot(s2[1,], s2[2,], col = alpha("blue", .3), pch = 16,
     main = "Laplace (original sources)", xlab = "s1", ylab = "s2")
hist(s2[1,], main = "Laplace s1 original", xlab = "")
hist(s2[2,], main = "Laplace s2 original", xlab = "")

plot(proj2_w_rotated[1,], proj2_w_rotated[2,], col = alpha("blue", .3), pch = 16,
     main = "Laplace (whitened & rotated)", xlab = "s1", ylab = "s2")
hist(proj2_w_rotated[1,], main = "Laplace s1 whitened & rotated", xlab = "")
hist(proj2_w_rotated[2,], main = "Laplace s2 whitened & rotated", xlab = "")


plot(s3[1,], s3[2,], col = alpha("blue", .3), pch = 16,
     main = "Uniform (original sources)", xlab = "s1", ylab = "s2")
hist(s3[1,], main = "Uniform s1 original", xlab = "")
hist(s3[2,], main = "Uniform s2 original", xlab = "")

plot(proj3_w_rotated[1,], proj3_w_rotated[2,], col = alpha("blue", .3), pch = 16,
     main = "Uniform (whitened & rotated)", xlab = "s1", ylab = "s2")
hist(proj3_w_rotated[1,], main = "Uniform s1 whitened & rotated", xlab = "")
hist(proj3_w_rotated[2,], main = "Uniform s2 whitened & rotated", xlab = "")
```

• Compare the histograms after rotation by θmin and θmax for the different distributions.
```{r}
par(mfrow = c(2,2))

min1 = which.min(k_after_rotation1[,1])
min2 = which.min(k_after_rotation2[,1])
min3 = which.min(k_after_rotation3[,1])

R_min1 = rotation_matrix(thetas[min1])
R_max1 = rotation_matrix(max1)
R_min2 = rotation_matrix(thetas[min2])
R_max2 = rotation_matrix(max2)
R_min3 = rotation_matrix(thetas[min3])
R_max3 = rotation_matrix(max3)

proj1_w_rotated_max = R_max1 %*% t(proj1_w)
proj2_w_rotated_max = R_max2 %*% t(proj2_w)
proj3_w_rotated_max = R_max3 %*% t(proj3_w)

proj1_w_rotated_min = R_min1 %*% t(proj1_w)
proj2_w_rotated_min = R_min2 %*% t(proj2_w)
proj3_w_rotated_min = R_min3 %*% t(proj3_w)

hist(proj1_w_rotated_max[1,], main = "Normal s1 theta_max", xlab = "")
hist(proj1_w_rotated_max[2,], main = "Normal s2 theta_max", xlab = "")
hist(proj1_w_rotated_min[1,], main = "Normal s1 theta_min", xlab = "")
hist(proj1_w_rotated_min[2,], main = "Normal s2 theta_min", xlab = "")

hist(proj2_w_rotated_max[1,], main = "Laplace s1 theta_max", xlab = "")
hist(proj2_w_rotated_max[2,], main = "Laplace s2 theta_max", xlab = "")
hist(proj2_w_rotated_min[1,], main = "Laplace s1 theta_min", xlab = "")
hist(proj2_w_rotated_min[2,], main = "Laplace s2 theta_min", xlab = "")

hist(proj3_w_rotated_max[1,], main = "Uniform s1 theta_max", xlab = "")
hist(proj3_w_rotated_max[2,], main = "Uniform s2 theta_max", xlab = "")
hist(proj3_w_rotated_min[1,], main = "Uniform s1 theta_min", xlab = "")
hist(proj3_w_rotated_min[2,], main = "Uniform s2 theta_min", xlab = "")

```
