---
title: "hw6_ma"
output: pdf_document
---
setwd("/Users/maxand22/Google Drive/Humboldt/4. Semester/Machine Intelligence II/Machine_Intelligence_II/hw5")

```{r}
library(audio)

install.packages("VGAM")
library(VGAM)
```


6.1.a 
Extend your code from the previous problem sheet to get an ICA-learning scheme based on the natural gradient with a learning rate ε that decays slowly to 0 (e.g. εt+1 = λεt with λ ≈ 1, λ < 1). Note that depending on λ you have to iterate over the (shuffled) data more than once for proper convergence.

Initializsation of last Excersise
```{r}
# a)
s1 = read.table("sound1.dat", header=FALSE)
s2 = read.table("sound2.dat", header=FALSE)

S = t(as.matrix(data.frame(s1,s2)))

# b)
set.seed(1234)
A = matrix(runif(4, 0, 1), nrow = 2)

X = A%*%S


# e)
X[1,] = X[1,] - mean(X[1,])
X[2,] = X[2,] - mean(X[2,])
```


Extention by lambda decreasing to zero and Plot of S_head
(assumption lambda ist not the eigenvalue)
```{r}
# logistic function
f = function(x){
    return(1/(1+exp(-x)))
}


# natural Gradient
t = 1
eta_t = 0.7
alpha = 1
lambda = 0.9102 #0.9112
set.seed(9991)
W2 = matrix(runif(4, 0, 1), ncol = 2)


unmixing_natural = function(W, X, n_steps = 18000){
    
    for(t in 1:n_steps){
        eta_t = eta_t*lambda
        
        if(alpha > n_steps-12){
           print(eta_t)
        }
        
        x = X[,alpha]
        
        c <-  do.call(cbind, replicate(nrow(W), x, simplify=FALSE))
        
        f_wx = 1 - 2*f(W%*%c)
        
        wx = W%*%c
        
        k_delta = diag(nrow(W))
        
        
        W_delta = eta_t*((k_delta + f_wx%*%wx)%*%W)
        
        W = W + W_delta
        
        alpha = alpha + 1
        if(alpha == n_steps){
            alpha = 1
        }
    }
    return(W)
}


W_natural = unmixing_natural(W2, X)
S_hat_natural = W_natural%*%X

cor(t(S_hat_natural), t(S))
plot(1:18000, S_hat_natural[1,], type = 'l', main = 'unmixed bird', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_natural[2,], type = 'l', main = 'unmixed halleluja', xlab = 't', ylab = 'value')

#play(audioSample(t(as.matrix(S_hat_natural[1,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural[2,])), rate = 8192))

```


6.1.b
Use the two sound signals from the last problem sheet and add (as third source s3) an ad- ditional “noise” source (normally distributed random numbers with a standard deviation similar to the two signals). Mix the signals using a mixing matrix of your choice and apply your ICA-algorithm. Plot the Mixed Sounds and recovered Sources


Plot initialised the third source normal distributed
```{r}

set.seed(222)
s3 <- t(matrix(rnorm(nrow(s1),0,((sd(as.matrix(s1)) + sd(as.matrix(s2)))/2)),nrow = 1))
S3 = t(as.matrix(data.frame(s1,s2,s3)))

plot(1:18000, s3, type = 'l', main = 'source 3 normal distributed', xlab = 't', ylab = 'value')

```

plot the created observations X
```{r}

set.seed(1234)
A3 = matrix(runif(9, 0, 1), nrow = 3)
X3 = A3%*%S3

#centering
X3[1,] = X3[1,] - mean(X3[1,])
X3[2,] = X3[2,] - mean(X3[2,])
X3[3,] = X3[3,] - mean(X3[3,])

plot(1:18000, X3[1,], type = 'l', main = 'observation 1', xlab = 't', ylab = 'value')
plot(1:18000, X3[2,], type = 'l', main = 'observation 2', xlab = 't', ylab = 'value')
plot(1:18000, X3[3,], type = 'l', main = 'observation 3', xlab = 't', ylab = 'value')


```


Plot the unmixed Xb by modified natural gradient 3x3
```{r}
# natural Gradient
t = 1
eta_t = 0.3
alpha = 1
lambda = 0.9102 #0.9102
set.seed(9991)
W3 = matrix(runif(9, 0, 1), ncol = 3)

W_natural3 = unmixing_natural(W3, X3)
S_hat_natural3 = W_natural3%*%X3

cor(t(S_hat_natural3), t(S3))
plot(1:18000, S_hat_natural3[1,], type = 'l', main = 's_hat bird', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_natural3[2,], type = 'l', main = 's_hat halleluja', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_natural3[3,], type = 'l', main = 's_hat source 3 normal distributed', xlab = 't', ylab = 'value')
#play(audioSample(t(as.matrix(S_hat_natural3[1,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[2,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[3,])), rate = 8192))
```


(c) Do the same analysis but adding a different “noise”-source (e.g. Laplace distributed) instead of the normal one.

Plot initialised the third source t distributed
```{r}

set.seed(666666)
st <- t(matrix(rt(nrow(s1),10),nrow = 1))
St = t(as.matrix(data.frame(s1,s2,st)))

plot(1:18000, st, type = 'l', main = 'source 3 t distributed', xlab = 't', ylab = 'value')

```

plot the created observations X laplace distributed
```{r}

set.seed(1234)
A3 = matrix(runif(9, 0, 1), nrow = 3)
Xt = A3%*%St

Xt[1,] = Xt[1,] - mean(Xt[1,])
Xt[2,] = Xt[2,] - mean(Xt[2,])
Xt[3,] = Xt[3,] - mean(Xt[3,])


plot(1:18000, Xt[1,], type = 'l', main = 'observation 1', xlab = 't', ylab = 'value')
plot(1:18000, Xt[2,], type = 'l', main = 'observation 2', xlab = 't', ylab = 'value')
plot(1:18000, Xt[3,], type = 'l', main = 'observation 3', xlab = 't', ylab = 'value')


```

plot S_hat noise source laplace distrubuted
```{r}
# natural Gradient
t = 1
eta_t = 0.25
alpha = 1
lambda = 0.91 #0.9102
set.seed(9991)
Wt = matrix(runif(9, 0, 1), ncol = 3)

W_naturalt = unmixing_natural(Wt, Xt)
S_hat_naturalt = W_naturalt%*%Xt

cor(t(S_hat_natural3), t(S3))
plot(1:18000, S_hat_naturalt[1,], type = 'l', main = 's_hat bird', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_naturalt[2,], type = 'l', main = 's_hat halleluja', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_naturalt[3,], type = 'l', main = 's_hat source 3 t distributed', xlab = 't', ylab = 'value')
#play(audioSample(t(as.matrix(S_hat_natural3[1,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[2,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[3,])), rate = 8192))
```




6.2 Moments of univariate distributions

Calculate the first 4 moments of the different random variables depending on the respective pa- rameters. In addition to providing the derivation (e.g. by using the characteristic function) fill the following table:

Laplace Distriution
```{r}
M = matrix(1, ncol = 3, nrow = 4)

b <- 1
mean <- 0

S_l <- rlaplace(nrow(s1),mean,b)

n = 1

L_X <- function(z){
  l_x <- (function(x){
          y <- ((1/(2*b))*exp((-1*abs(x-mean))/b))*(x^n)
          })
  return(l_x)
}

I_L_X <- function(z){integrate(L_X(z), lower = -100, upper = 100)$value}

for(i in 1:nrow(M)){
  n<-i
  M[i,1] <- I_L_X(S_l)
}

M[,1]

```

Normal Distribution
Create Matrix and function for moment calculation of normal distributon
+
plot Moments for normal distribution and s3
```{r}

pi = 22/7
sd = ((sd(as.matrix(s1)) + sd(as.matrix(s2)))/2)
mean = 0

n = 1

P_X <- function(z){
  p_x <- (function(x){
          y <- (exp((-1*(x-mean)^2)/(2*sd*sd))/((2*pi*sd)^0.5))*(x^n)
          })
  return(p_x)
}

I_P_X <- function(z){integrate(P_X(z), lower = -100, upper = 100)$value}

for(i in 1:nrow(M)){
  n<-i
  M[i,2] <- I_P_X(s3)
}

M[,2]

```

Uniform Distribution Moments
with
Probility density function depending on a(min) and b(max)
```{r}

a = 0
b = 1
S_uni <- runif(nrow(s1),a,b)


n = 1

U_X <- function(z){
  u_x <- (function(x){
          y <- (1/(b-a))*(x^n)
          })
  return(u_x)
}

I_U_X <- function(z){integrate(U_X(z), lower = a, upper = b)$value}

for(i in 1:nrow(M)){
  n<-i
  M[i,3] <- I_U_X(S_uni)
}

M[,3]

```
