---
title: "HW6_MI2"
author: "Benjamin Kolb"
date: "13 Juni 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Benni/Box Sync/_studium/2. Semester/04_Machine Intelligence II/Exercises/06")
knitr::opts_chunk$set(error = TRUE)
#load("test.RData")
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

#https://stackoverflow.com/questions/11155182/is-there-a-way-to-knitr-markdown-straight-out-of-your-workspace-using-rstudio
```

Packages used
```{r results='hide', message=FALSE, warning=FALSE}
library(audio)
library(VGAM)
```

Exercise 6.1: Natural Gradient

(a)
Initializsation of last Excersise
```{r}
s1 = read.table("sound1.dat", header=FALSE)
s2 = read.table("sound2.dat", header=FALSE)

S = t(as.matrix(data.frame(s1,s2)))

set.seed(1234)
A = matrix(runif(4, 0, 1), nrow = 2)

X = A%*%S

X[1,] = X[1,] - mean(X[1,])
X[2,] = X[2,] - mean(X[2,])
```

Extention by lambda decreasing to zero and Plot of S_head
```{r}
# logistic function
f = function(x){
    return(1/(1+exp(-x)))
}


# natural Gradient
t = 1
eta_t = 0.7
alpha = 1
lambda = 0.9102 #0.9112
set.seed(9991)
W2 = matrix(runif(4, 0, 1), ncol = 2)


unmixing_natural = function(W, X, n_steps = 18000){
    
    for(t in 1:n_steps){
        eta_t = eta_t*lambda
        
        if(alpha > n_steps-12){
           print(eta_t)
        }
        
        x = X[,alpha]
        
        c <-  do.call(cbind, replicate(nrow(W), x, simplify=FALSE))
        
        f_wx = 1 - 2*f(W%*%c)
        
        wx = W%*%c
        
        k_delta = diag(nrow(W))
        
        
        W_delta = eta_t*((k_delta + f_wx%*%wx)%*%W)
        
        W = W + W_delta
        
        alpha = alpha + 1
        if(alpha == n_steps){
            alpha = 1
        }
    }
    return(W)
}


W_natural = unmixing_natural(W2, X)
S_hat_natural = W_natural%*%X

cor(t(S_hat_natural), t(S))
plot(1:18000, S_hat_natural[1,], 
     type = 'l', 
     main = 'unmixed bird', 
     xlab = 't', 
     ylab = 'value')
plot(1:18000, S_hat_natural[2,], 
     type = 'l', 
     main = 'unmixed halleluja', 
     xlab = 't', 
     ylab = 'value')

```

(b)
Plot initialised the third source normal distributed
```{r}
set.seed(222)
S_uni <- t(matrix(rnorm(nrow(s1),0,((sd(as.matrix(s1)) + sd(as.matrix(s2)))/2)),nrow = 1))
S_uni = t(as.matrix(data.frame(s1,s2,S_uni)))

plot(1:18000, S_uni, type = 'l', main = 'source 3 normal distributed', xlab = 't', ylab = 'value')
```

plot the created observations X
```{r}

set.seed(1234)
A3 = matrix(runif(9, 0, 1), nrow = 3)
X3 = A3%*%S_uni

#centering
X3[1,] = X3[1,] - mean(X3[1,])
X3[2,] = X3[2,] - mean(X3[2,])
X3[3,] = X3[3,] - mean(X3[3,])

plot(1:18000, X3[1,], type = 'l', main = 'observation 1', xlab = 't', ylab = 'value')
plot(1:18000, X3[2,], type = 'l', main = 'observation 2', xlab = 't', ylab = 'value')
plot(1:18000, X3[3,], type = 'l', main = 'observation 3', xlab = 't', ylab = 'value')


```


Plot the unmixed Xb by modified natural gradient 3x3
```{r}
# natural Gradient
t = 1
eta_t = 0.3
alpha = 1
lambda = 0.9102 #0.9102
set.seed(9991)
W3 = matrix(runif(9, 0, 1), ncol = 3)

W_natural3 = unmixing_natural(W3, X3)
S_hat_natural3 = W_natural3%*%X3

cor(t(S_hat_natural3), t(S_uni))
plot(1:18000, S_hat_natural3[1,], type = 'l', main = 's_hat bird', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_natural3[2,], type = 'l', main = 's_hat halleluja', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_natural3[3,], type = 'l', main = 's_hat source 3 normal distributed', xlab = 't', ylab = 'value')
#play(audioSample(t(as.matrix(S_hat_natural3[1,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[2,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[3,])), rate = 8192))
```

(c)
```{r}

set.seed(666666)
st <- t(matrix(rt(nrow(s1),10),nrow = 1))
St = t(as.matrix(data.frame(s1,s2,st)))

plot(1:18000, st, type = 'l', main = 'source 3 t distributed', xlab = 't', ylab = 'value')

```

```{r}

set.seed(1234)
A3 = matrix(runif(9, 0, 1), nrow = 3)
Xt = A3%*%St

Xt[1,] = Xt[1,] - mean(Xt[1,])
Xt[2,] = Xt[2,] - mean(Xt[2,])
Xt[3,] = Xt[3,] - mean(Xt[3,])


plot(1:18000, Xt[1,], type = 'l', main = 'observation 1', xlab = 't', ylab = 'value')
plot(1:18000, Xt[2,], type = 'l', main = 'observation 2', xlab = 't', ylab = 'value')
plot(1:18000, Xt[3,], type = 'l', main = 'observation 3', xlab = 't', ylab = 'value')


```

plot S_hat noise source laplace distrubuted
```{r}
# natural Gradient
t = 1
eta_t = 0.25
alpha = 1
lambda = 0.91 #0.9102
set.seed(9991)
Wt = matrix(runif(9, 0, 1), ncol = 3)

W_naturalt = unmixing_natural(Wt, Xt)
S_hat_naturalt = W_naturalt%*%Xt

cor(t(S_hat_natural3), t(S_uni))
plot(1:18000, S_hat_naturalt[1,], type = 'l', main = 's_hat bird', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_naturalt[2,], type = 'l', main = 's_hat halleluja', xlab = 't', ylab = 'value')
plot(1:18000, S_hat_naturalt[3,], type = 'l', main = 's_hat source 3 t distributed', xlab = 't', ylab = 'value')
#play(audioSample(t(as.matrix(S_hat_natural3[1,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[2,])), rate = 8192))
#play(audioSample(t(as.matrix(S_hat_natural3[3,])), rate = 8192))
```

6.2: Moments of univariate distributions
Laplace Distriution
```{r}
M = matrix(1, ncol = 3, nrow = 4)

b <- 1
mean <- 0

S_l <- rlaplace(nrow(s1),mean,b)

m1 <- mean(S_l)
m2 <- sum((S_l-m1)^2)/length(S_l)
m3 <- (sum((S_l-m1)^3)/18000)/((sum((S_l-m1)^2)/18000)^(3/2))
(sum(S_l-mean)^3/18000)
m4 <- (sum((S_l-mean)^4)/18000)/((sum((S_l-mean)^2)/18000)^(4/2))

var(S_l)
skewness(S_l)
kurtosis(S_l)

m1L <- mean(S_l)
m2L <- sum((S_l-m1)^2)/length(S_l)
m3L <- (sum((S_l-m1)^3)/18000)/((sum((S_l-m1)^2)/18000)^(3/2))
m4L <- (sum((S_l-m1)^4)/18000)/((sum((S_l-m1)^2)/18000)^(4/2))

```


Normal Distribution
Create Matrix and function for moment calculation of normal distributon
+
plot Moments for normal distribution and S_uni
```{r}

pi = 22/7
sd = ((sd(as.matrix(s1)) + sd(as.matrix(s2)))/2)
mean = 0

n = 1

P_X <- function(z){
  p_x <- (function(x){
          y <- (exp((-1*(x-mean)^2)/(2*sd*sd))/((2*pi*sd)^0.5))*(x^n)
          })
  return(p_x)
}

I_P_X <- function(z){integrate(P_X(z), lower = -100, upper = 100)$value}

for(i in 1:nrow(M)){
  n<-i
  M[i,2] <- I_P_X(S_uni)
}

M[,2]

#S_uni

m1n <- mean(S_uni)
m2n <- sum((S_uni-m1n)^2)/length(S_uni)
m3n <- (sum((S_uni-m1n)^3)/18000)/((sum((S_uni-m1n)^2)/18000)^(3/2))
m4n <- (sum((S_uni-m1n)^4)/18000)/((sum((S_uni-m1n)^2)/18000)^(4/2))

mean(S_uni)
var(S_uni)
skewness(S_uni)
kurtosis(S_uni)

```

Uniform Distribution Moments
with
Probility density function depending on a(min) and b(max)
```{r}

a = 0
b = 1
S_uni <- runif(nrow(s1),a,b)


n = 1

U_X <- function(z){
  u_x <- (function(x){
          y <- (1/(b-a))*(x^n)
          })
  return(u_x)
}

I_U_X <- function(z){integrate(U_X(z), lower = a, upper = b)$value}

for(i in 1:nrow(M)){
  n<-i
  M[i,3] <- I_U_X(S_uni)
}

m1u <- mean(S_uni)
m2u <- sum((S_uni-m1u)^2)/length(S_uni)
m3u <- (sum((S_uni-m1u)^3)/18000)/((sum((S_uni-m1u)^2)/18000)^(3/2))
m4u <- (sum((S_uni-m1u)^4)/18000)/((sum((S_uni-m1u)^2)/18000)^(4/2))

mean(S_uni)
var(S_uni)
skewness(S_uni)
kurtosis(S_uni)

M[,3]

```