---
title: "mi2_hw2"
author: "Nikolas Höft"
date: "26 4 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Machine Intelligence 2 Homework 1



```{r, warning=FALSE, message=FALSE, error=FALSE}
# load packages
#install.packages("imager")
library(imager)

#install.packages("foreign")
library(foreign)

#install.packages("gridExtra")
library(gridExtra)

#install.packages("ggplot2")
library(ggplot2)

#install.packages("reshape2")
library(reshape2)

#install.packages("colorRamps")
library(colorRamps)

#setwd("~/Desktop/Uni/Statistik Master/courses/machine intelligence 2/data/natIMG.jpeg")

```

```{r}
# 2.1 a)

# load data
pcaData2d <- read.csv("Ex2/pca-data-2d.txt", sep = "", header = FALSE)

# calcualte mean matrix
n <- nrow(pcaData2d)
M_mean <- matrix(data=1, nrow=n) %*% cbind(mean(pcaData2d[,1]),
                                           mean(pcaData2d[,2]))

# subtract mean from data set -> "Difference Matrix"
pcaData2d_cen <- pcaData2d - M_mean

# plot centered matrix
ggplot(pcaData2d_cen, aes(x=pcaData2d_cen$V1, y=pcaData2d_cen$V2)) + geom_point()
```


```{r}
# 2.1 b)

# convert df to matrix
D <- as.matrix(pcaData2d_cen)

# compute Covariance Matrix C
C <- (n-1)^-1 * t(D) %*% D

# Eigenvalues of C, Eigenvectors in P
ev <- eigen(C)
P  <- ev$vectors

# transform D
D_tr <- P %*% t(D)

# in ggplot2 plottable data frame
X_pca <- as.data.frame(t(D_tr)) 

# plot of the transformed set
ggplot(X_pca, aes(x=X_pca$V1, y=X_pca$V2)) + geom_point()
```

```{r}
# 2.1 c)


# plot only PC1 
ggplot(X_pca, aes(x=X_pca$V1, y=0)) + geom_point()

# plot only PC2
ggplot(X_pca, aes(x=0, y= X_pca$V2)) + geom_point()
```

```{r}
# 2.2 a)


# load data set
pcaData3d <- read.csv("Ex2/pca-data-3d.txt")

# centering
n_3 <- nrow(pcaData3d)
M3_mean <- matrix(data=1, nrow=n_3) %*% colMeans(pcaData3d)

# subtract mean from data set -> "Difference Matrix"
D3 <- pcaData3d - M3_mean

# scatterplot matrix
pairs(D3)
```

```{r}
# 2.2 b)

# convert df to matrix
D3 <- as.matrix(D3)

# compute Covariance Matrix C
C3 <- (n_3-1)^-1 * t(D3) %*% D3

# Eigenvalues of C3, Eigenvectors in P3
ev3 <- eigen(C3)
P3  <- ev3$vectors

# transform D3
D3 <- P3 %*% t(D3)
X_3pca <- as.data.frame(t(D3)) 
pairs(X_3pca)
```

```{r}
# 2.2 c)


D3 <- pcaData3d - M3_mean

# convert df to matrix
D3 <- as.matrix(D3)

# compute Covariance Matrix C
C3 <- (n_3-1)^-1 * t(D3) %*% D3

eigen_vals <- eigen(C3)$values
eigen_vecs <- eigen(C3)$vectors

# compute pc scores
#pc_scores <- prcomp(D3)

pc_scores_1 <- D3 %*% eigen_vecs[, 1]
pc_scores_1_2 <- D3 %*% eigen_vecs[, 1:2]
pc_scores_1_2_3 <- D3 %*% eigen_vecs[, 1:3]


pca_reconstruction_1 <- pc_scores_1 %*% t(eigen_vecs[, 1])
pairs(pca_reconstruction_1, main = "PCs for reconstruction: 1")

pca_reconstruction_1_2 <- pc_scores_1_2 %*% t(eigen_vecs[, 1:2])
pairs(pca_reconstruction_1_2, main = "PCs for reconstruction: 1, 2")

pca_reconstruction_1_2_3 <- pc_scores_1_2_3 %*% t(eigen_vecs[, 1:3])
pairs(pca_reconstruction_1_2_3, main = "PCs for reconstruction: 1, 2, 3")

# for reconstruction of uncentered data, add mean veactor to each column (add M3)
```

** Interpretation**
Obviously, the first two PCs are useful here. Plotting only the first PC alone 
does not yield a lot of information, only about the spread in this direction. 
When taking into account the first two PCs, one can see the (possibly) sinusoidal
trend in the data. Using all three PCs yields the original centered COV matrix
that was plotted in 2.2 a). This is oubvious because due to the orthonormality
of the eigenvectors eigen_vecs %*% t(eigen_vecs) will yield an identity matrix,
hence, when multiplying this matrix with the centered COV matrix D3, the COV
matrix remains unchanged.


```{r}
# 2.3 a)

# load data
expDat <- read.csv("Ex2/expDat.txt", sep = ",", header = TRUE)

# calculate mean matrix
n <- nrow(expDat[2:ncol(expDat)])
M_mean <- matrix(data=1, nrow=n) %*% colMeans(expDat[2:ncol(expDat)])
# subtract mean from data set 
expDat_cen <- expDat[2:ncol(expDat)] - M_mean

# convert df to matrix
D <- as.matrix(expDat_cen)
# compute Covariance Matrix C
C <- (n-1)^-1 * t(D) %*% D
# Eigenvalues of C, Eigenvectors in P
ev <- eigen(C)
P  <- ev$vectors

```



```{r}
# 2.3 b)

# transform D
D_tr <- P %*% t(D)
# in ggplot2 plottable data frame
X_pca <- as.data.frame(t(D_tr)) 

#scatter plot
ggplot(X_pca, aes(x=X_pca$V1, y=X_pca$V2)) + scale_colour_gradient(low = "red", high = "blue") + geom_point(aes(colour=1:100))


#line plot
df <- data.frame(expDat[,1], X_pca$V1, X_pca$V2)
colnames(df)[1] <- "time_index"

ggplot(df, aes(x=time_index, y=value)) + 
    scale_colour_gradient(low = "red", high = "blue") + 
    geom_line(aes(y=X_pca$V1, colour=1:100)) + 
    geom_line(aes(y=X_pca$V2,colour=1:100))


```

```{r}
# 2.3 c)

newDat <- apply(expDat[2:21], 2, sample)
```

```{r}
# 2.3 d)

# shuffled dataset
# calculate mean matrix of new data set
n_new <- nrow(newDat)
M_mean_new <- matrix(data=1, nrow=n_new) %*% colMeans(newDat)
# subtract new mean from shuffled data set
Dat_cen_new <- newDat - M_mean_new
# compute Covariance Matrix C_new 
C_new <- (n_new-1)^-1 * t(Dat_cen_new) %*% Dat_cen_new
```

**2.3 e)** 

Shuffeling the data rowwise in the same sequence for all columns does not affect
the resulting covariance matrix. This is easy to see without programming, since
both, variances and covariances are based on element-wise sums of squared errors. 
For summing over a number of elements, the order is not relevant. Therefore, the 
variances remain unchanged even if the data was not shuffeled in the same sequence
since they do not depend on the values of other variables.
However, for the covariances this is important because otherwise the relationship
between the features would change. But here, since the rows 
are shuffeled in the same way for all columns (i.e. features), the relationships 
(covariances) between the features remain the same as well.


```{r}
# 2.4 a)

```

